An Efficient Minibatch Acceptance Test for MCMC

### 2016-06-21 nips_2016.tex
correct typos

\begin{itemize}
\item We develop a new minibatch acceptance test which satisfies detailed balance.
\item we developed effectively measurable bounds for both absolute and relative
  test error. 
\item We compare performance of our new test and prior approaches. 
\item We experiment using the approach for posterior estimation.
\end{itemize}

@@

\begin{proof} see appendix
  \end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more agressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

	\item If the ${\rm std}(\Delta') > 1.2$, the decomposition of $X$ (i.e. Equation~\ref{eq:deconvolution}) is unstable. We consume more minibatches to decrease the ${\rm std}(\Delta')$. We update the value of $\Delta'$ by $\Delta' = (k\Delta'_{\rm old} + \Delta'_{\rm new}) / (k+1)$, where $\Delta'_{\rm old}$ is the existing $\Delta'$ which uses $k$ minibatches to computed (usually $k=1$), and $\Delta'_{\rm new}$ is the $\Delta'$ computed from the new minibatch. We also update the value of ${\rm std}(\Delta')$ by ${\rm std}(\Delta') = \sqrt{\left ( k^2 {\rm var}(\Delta'_{\rm old}) + {\rm var}(\Delta'_{\rm new}) \right ) / (k+1)^2}$. Thus, we can decrease the ${\rm std}(\Delta')$ by consuming more data until the ${\rm std}(\Delta') > 1.2$. With help of temperature (see Section~\ref{ssec:discussion}), we will show that usually the mean of minibatch consumed per iteration is below 2 in Section~\ref{sec:experiments}. Later our efficiency analysis in Section~\ref{sec:theory} will reveal that the number of data instance consumed by our algorithm is much fewer than that of exisitng algorithms.

@@

\begin{theorem}\label{thm:efficiency}
We require $\frac{ C^2 \rho^2 }{\sigma^6 \varepsilon^2}$ minibatch size to ensure $|P_a -P_a'| < \varepsilon$, where $C$ is a positive constant which is smaller than 0.4748, $\sigma^2 $ and $\rho$ is the second and third moment of $(\nabla\log \Pr(x_i|\theta_t) - \nabla\log\Pr(x_i|\theta')) - E\left[ \nabla\log \Pr(x_i|\theta_t) - \nabla\log\Pr(x_i|\theta')\right]$, $P_a$ is the accpet rate from the complete data, $P_a'$ is the accpet rate from minibatch data by our algorithm, $\theta_t$ is the current parameter, and $\theta'$ is the proposed parameter.

@@

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$ samples for the four
methods, with darker regions indicating a greater density of points. The first three methods obtain the
same rough form of the posterior, so our MH test can indeed result in the same posterior as the
other two methods. However, the fourth method (i.e. just treat the $\Delta'$ which comes from minibatch as $\Delta$ which comes from the complete data) reveals wider spread distribution than the posterior. From Figure~\ref{fig:gauss_mix_1} we illustrate our proposed method can obtain the samples which have almost the same distirbution as the standard MH test from the complete data, adaptive minibatch test, and the true posterior. Meanwhile, we show the benefits of our minibatch method compared with the navie minibatch method.

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (on a log scale for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in lots of wasted computation.  For this particular experiment, that happened eight times. The mean of adaptive minibatch size is 520 which is roughly close to $\sqrt{N}$ (i.e. 100). This number validates the claims of Theorem~\ref{thm:efficiency}. In
contrast, our method keeps the minibatch size around 200. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the minibatch size, and the mean of our minibatch size is 247. 
The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
\ge 1.2$ for only consuming one mimnibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
our MH test five times and averaged the number of ``violations.'' As the size increases, the
number of violations decreases, as expected.

Figure~\ref{fig:gauss_mix_3} (to the left) shows a historgram of ${\rm std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows the confidence interval of our estimated ${\rm std}(\Delta')$ at each iteration \footnote{In order to show the confidence interval clearly, we use moving smooth to smooth the confidence interval, and plot each value by interval of 50 iterations}. The blue line represents the standard deviation comes from the complete data, which can be treated as the ``true'' ${\rm std}(\Delta')$, the green line represents our estimation of ${\rm std}(\Delta')$ from the minibatch, and the two red dash lines represent the confidence interval of ${\rm std}(\Delta')$ by adding / subtracking one ${\rm std(std(}\Delta'))$. This figure indicates that our estimation of ${\rm std}(\Delta')$ from one minibatch data is accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value of $\rho/\sigma^3$ at each iteration. Recall Theorem~\ref{thm:efficiency}, the number of data instance required is $\frac{4 \ell^2 C^2 \rho^2 }{\sigma^6 \varepsilon^2}$. This figure reveals that $\rho/\sigma^3$ is rougly a constant, which means the data required for each iteration is a constant value which has nothing to do with the $N$.

@@

If the minibatch size is $n$, we fisrt compute the probability that more than $n$ data instances are required for adaptive sampling method for one iteration. We denote this probaility $P_{rej}$.
\[P_{rej} = \Pr \left ( 1-\phi_{n-1} \left( \frac{|\bar{l} - \frac{1}{N} \log (u\frac{p(\theta_t)q(\theta'|\theta_t)}{p(\theta')q(\theta_t|\theta')})|}{s} \right )  \geq \varepsilon \right ) \]

>>>>

\begin{itemize}
\item We developed a new minibatch acceptance test which satisfies detailed balance.
\item we developed effectively measurable bounds for both absolute and relative
  test error. 
\item We compared performance of our new test and prior approaches. 
\item We did experiments using the approach for posterior estimation.
\end{itemize}

@@

\begin{proof} see appendix
  \end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

	\item If the ${\rm std}(\Delta') > 1.2$, the decomposition of $X$ (i.e. Equation~\ref{eq:deconvolution}) is unstable. We consume more minibatches to decrease the ${\rm std}(\Delta')$. We update the value of $\Delta'$ by $\Delta' = (k\Delta'_{\rm old} + \Delta'_{\rm new}) / (k+1)$, where $\Delta'_{\rm old}$ is the existing $\Delta'$ which uses $k$ minibatches to computed (usually $k=1$), and $\Delta'_{\rm new}$ is the $\Delta'$ computed from the new minibatch. We also update the value of ${\rm std}(\Delta')$ by ${\rm std}(\Delta') = \sqrt{\left ( k^2 {\rm var}(\Delta'_{\rm old}) + {\rm var}(\Delta'_{\rm new}) \right ) / (k+1)^2}$. Thus, we can decrease the ${\rm std}(\Delta')$ by consuming more data until the ${\rm std}(\Delta') > 1.2$. With help of temperature (see Section~\ref{ssec:discussion}), we will show that usually the mean of minibatch consumed per iteration is below 2 in Section~\ref{sec:experiments}. Later our efficiency analysis in Section~\ref{sec:theory} will reveal that the number of data instance consumed by our algorithm is much fewer than that of existing algorithms.

@@

\begin{theorem}\label{thm:efficiency}
We require $\frac{ C^2 \rho^2 }{\sigma^6 \varepsilon^2}$ minibatch size to ensure $|P_a -P_a'| < \varepsilon$, where $C$ is a positive constant which is smaller than 0.4748, $\sigma^2 $ and $\rho$ is the second and third moment of $(\nabla\log \Pr(x_i|\theta_t) - \nabla\log\Pr(x_i|\theta')) - E\left[ \nabla\log \Pr(x_i|\theta_t) - \nabla\log\Pr(x_i|\theta')\right]$, $P_a$ is the accept rate from the complete data, $P_a'$ is the accept rate from minibatch data by our algorithm, $\theta_t$ is the current parameter, and $\theta'$ is the proposed parameter.

@@

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$ samples for the four
methods, with darker regions indicating a greater density of points. The first three methods obtain the
same rough form of the posterior, so our MH test can indeed result in the same posterior as the
other two methods. However, the fourth method (i.e. just treat the $\Delta'$ which comes from minibatch as $\Delta$ which comes from the complete data) reveals wider spread distribution than the posterior. From Figure~\ref{fig:gauss_mix_1} we illustrate our proposed method can obtain the samples which have almost the same distribution as the standard MH test from the complete data, adaptive minibatch test, and the true posterior. Meanwhile, we show the benefits of our minibatch method compared with the naive minibatch method.

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (on a log scale for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in lots of wasted computation.  For this particular experiment, that happened eight times. The mean of adaptive minibatch size is 520 which is roughly close to $\sqrt{N}$ (i.e. 100). This number validates the claims of Theorem~\ref{thm:efficiency}. In
contrast, our method keeps the minibatch size around 200. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the minibatch size, and the mean of our minibatch size is 247. 
The third plot in Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta')
\ge 1.2$ for only consuming one minibatch data. For six minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with
our MH test five times and averaged the number of ``violations.'' As the size increases, the
number of violations decreases, as expected.

Figure~\ref{fig:gauss_mix_3} (to the left) shows a histogram of ${\rm std}(\Delta')$. Figure~\ref{fig:gauss_mix_3} (to the middle) shows the confidence interval of our estimated ${\rm std}(\Delta')$ at each iteration \footnote{In order to show the confidence interval clearly, we use moving smooth to smooth the confidence interval, and plot each value by interval of 50 iterations}. The blue line represents the standard deviation comes from the complete data, which can be treated as the ``true'' ${\rm std}(\Delta')$, the green line represents our estimation of ${\rm std}(\Delta')$ from the minibatch, and the two red dash lines represent the confidence interval of ${\rm std}(\Delta')$ by adding / subtracting one ${\rm std(std(}\Delta'))$. This figure indicates that our estimation of ${\rm std}(\Delta')$ from one minibatch data is accurate. Figure~\ref{fig:gauss_mix_3} (to the right) shows the value of $\rho/\sigma^3$ at each iteration. Recall Theorem~\ref{thm:efficiency}, the number of data instance required is $\frac{4 \ell^2 C^2 \rho^2 }{\sigma^6 \varepsilon^2}$. This figure reveals that $\rho/\sigma^3$ is roughly a constant, which means the data required for each iteration is a constant value which has nothing to do with the $N$.

@@

If the minibatch size is $n$, we first compute the probability that more than $n$ data instances are required for adaptive sampling method for one iteration. We denote this probability $P_{rej}$.
\[P_{rej} = \Pr \left ( 1-\phi_{n-1} \left( \frac{|\bar{l} - \frac{1}{N} \log (u\frac{p(\theta_t)q(\theta'|\theta_t)}{p(\theta')q(\theta_t|\theta')})|}{s} \right )  \geq \varepsilon \right ) \]
